An introduction to Docker for reproducible research, with examples from the R environment

### 2014-09-18 manuscript.Rmd
writing

Though Docker is not immune to these challenges, it offers a interesting 
example of a way forward in addressing these fundamental concerns.

- *Local environment*
- *Modular reuse*

@@

### Portable computation

A particular advantage of this approach is that the resulting
computational environment is immediately _portable_.  This is useful not
only for the purposes of reproducible research, where other users may seek
to reconstruct the computational environment necessary to run the code,
but is also of immediate value to the researcher themselves. For instance,
a researcher might want to execute their code on a cloud server which
has more memory or processing power then their local machine, or would
want a co-author to help debug a particular problem.  In either case,
the researcher can export a snapshot of their running container:

```bash
docker export <container-name> > container.tar
```
and then run this identical environment on the cloud or collaborators' machine.

## Docker remix: Building re-usable modules

@@

The key point here is that other researchers can easily build off this
image just created, extending our work directly, rather than having
to go back to the original image. Before others can re-use our image,
it must be made available. Docker provides a convenient way to do this
using the Docker Hub.

### Docker Hub

While Docker images tend to be much smaller than equivalent virtual
machines, moving around even 100's of gigabytes can be a challenge.
In their work on virtual machines, @Dudley_2010 recommend only running
these tools on cloud servers such as the Amazon EC2 system. As most
researchers still develop software locally and may not have ready access
to these resources, such a requirement adds an additional barrier to
reuse.  Docker provides a convenient way to share any image publicly
or privately through the Hub after creating a free account.  For instance,
one can  share a public copy of the image just created by using the
`docker push` command, followed by the name of the image using the command:

Another user can now build directly on our image, rather than on the `ubuntu:latest`
image used as a base in the first Dockerfile.  To do so, one simply specifies a
different image in the `FROM` directive of the Dockerfile, followed by the lines
required to add any additional software required.

@@

----------------

<!--
The implementation of this would nonetheless depend on the individual 
researcher's willingness to break things into multiple containers, as
it is indeed possible to run everything in a monolithic container. 
This is not necessarily bad, as it simplifies the workflow (no need
to execute and link separate containers), and can be refactored into
multiple containers when appropriate. 
-->

<!--

##  Best practices

- Using Dockerfiles instead of containers
- Using appropriate, common base images
- Archiving tarball snapshots
- Adding tests/checks to the Dockerfile
- Sharing 
- Leveraging package managers appropriately to handle software dependencies

@@

-------------

# Further considerations

@@

<!-- Dynamic documents --> Dynamic documents are another example of an
approach that

approach [@Leisch_2002; @Peng_2011; @Xie_2013], which seek to embed
the code required to re-generate the results within the manuscript
itself. This process serves as a way to distribute computational
code with publications, to ensure that methodological descriptions,
parameter values, figures, and so forth remain synchronized with the code,
and to provide a way for other researchers to both tweak the code and
'execute' the paper to reproduce the results.  While these features
are very desirable from the perspective of reproducible research,
they inevitably increase the number of software dependencies involved,
including those tools that support combining code with text [e.g. @Xie_2013]
as well as tools merely for the handling the formating and layout
(e.g. LaTeX, pandoc).  Consequently, such a paper can fail to execute
successfully for reasons having nothing to do with the computations
involved in the research, but rather due to software responsible for
implementing these dynamic features.  This problem is largely resolved by
providing a consistent development environment, making the approaches
discussed here a natural complement to that of dynamic documents.
A similar argument can be made regarding the use of workflow systems,
which can introduce additional software dependencies even as they seek
to make it easier to work across them.

>>>>

Though Docker is not immune to these challenges, it offers a interesting 
example of a way forward in addressing these fundamental concerns. Here 
I highlight five of these features in turn:

- Local environment
- Modular reuse
- Portable environments 
- public repository for sharing
- Versioning

@@

## Portable computation & sharing

A particular advantage of this approach is that the resulting
computational environment is immediately _portable_. LXC containers
by themselves are unlikely to run in the same way, if at all, across
different machines, due to differences in networking, storage,
logging and so forth. Docker handles the packaging and execution of
a container so that it works identically across different machines,
while exposing the necessary interfaces for networking ports, volumes,
and so forth.  This is useful not only for the purposes of reproducible
research, where other users may seek to reconstruct the computational
environment necessary to run the code, but is also of immediate value
to the researcher themselves. For instance, a researcher might want to
execute their code on a cloud server which has more memory or processing
power then their local machine, or would want a co-author to help debug a
particular problem.  In either case, the researcher can export a snapshot
of their running container:

```bash
docker export <container-name> > container.tar
```

and then run this identical environment on the cloud or collaborators' machine.

Sharing these images is further facilitated by the Docker Hub technology. 
While Docker images tend to be much smaller than equivalent virtual
machines, moving around even 100's of gigabytes can be a challenge.
In their work on virtual machines, @Dudley_2010 recommend only running
these tools on cloud servers such as the Amazon EC2 system. As most
researchers still develop software locally and may not have ready access
to these resources, such a requirement adds an additional barrier to
reuse.  Docker provides a convenient way to share any image publicly
or privately through the Hub after creating a free account.  Docker Hub
is both a publicly available service and also available as 
a separate open source platform that can be deployed on private servers.

One can  share a public copy of the image just created by using the
`docker push` command, followed by the name of the image using the command:

If a Dockerfile is made available on a public code repository such as
[Github](https://github.com) or [Bitbucket](https://bitbucket.org),
the Hub can automatically build the image whenever a change is made to
the Dockerfile, making the `push` command unnecessary. A user can update
their local image using the `docker pull <imagename>`, which downloads
any changes that have since been made to the copy of the image on the Hub.

## Re-usable modules

@@

The key point here is that other researchers can easily build off this
image just created, extending our work directly, rather than having
to go back to the original image.

If this image is shared (either directly or through a Docker Hub),
another user can now build directly on our image, rather than on the
`ubuntu:latest` image used as a base in the first Dockerfile.  To do
so, one simply specifies a different image in the `FROM` directive of
the Dockerfile, followed by the lines required to add any additional
software required.

@@

# Conclusions

## Best Practices

The effectiveness of this approach for supporting reproducible research
nonetheless depends on how each of these features are adopted and 
implemented by individual researchers. I summarize a few of the choices

- Use docker containers during development

A key feature of the Docker approach is the ability to mimic as closely
as possible the current workflow and development practices of the user.
Code executing inside a container on a local machine can appear identical
to code running natively, but with the added benefit that one can simply
recreate or snapshot and share the entire computational environment with
a few simple commands.  This works best if researchers set up their computational
environment in a container from the outset of the project.

- Write Dockerfiles instead installing interactive sessions.

As we have noted already, Docker can be used in a purely interactive
manner to record and distribute changes to a computational environment.
However, the approach is most useful for reproducible research when
researchers begin by defining their environment explicitly in the DevOps
fashion by writing a Dockerfile.

- Adding tests or checks to the Dockerfile

Dockerfile commands need not be limited to installing software, but
can also include execution.  This can help verify that an image has 
build successfully with all the software necessary to run the 
research code of interest.

- Use and provide appropriate base images

Though Docker supports modular design, it remains up to the researchers
to take advantage of it. An appropriate workflow might involve one
Dockerfile that includes all the software dependencies a researcher
usually uses in the course of their development, which can then be
extended by separate docker images for particular projects.  Re-using
existing images reduces the effort required to set up an environment,
contributes to the standardization of computational environments within
a field, and best leverages the ability of Docker's distribution system
to download only differences.

- Share Docker images and Dockerfiles

The Docker Hub significantly reduces the barriers for making even
large images (which can exceed the file size limits of journals common
scientific data repositories such as Dryad and Figshare) readily available
to other researchers.

- Archive tarball snapshots

Despite similar semantics to git, Docker's versioning system works rather 
differently than version management of code.  Docker can roll back AUFS
layers that have been added to an image, but not revert to the earlier 
state of a particular layer.  In consequence, to preserve a bitwise identical
snapshot of a container used to generate a given set of results, it is
necessary to archive the image tarball itself -- one can not simply
rely on the docker history to recover an earlier state.

## Limitations and future developments

Docker has the potential to address shortcomings of certain existing
approaches to reproducible research challenges that stem from recreating
complex computational environments.  Docker also provides a promising
case study in other issues. Its versioning, modular design, portable
containers, and simple interface have proven successful in industry and
could have promising implications for reproducible research in scientific
communities. Nonetheless, these advances raise questions and challenges 
of their own.

Docker does not provide complete virtualization but relies on the 
Linux kernel provided by the host.  Systems research can provide insight
on what limitations to reproducibility this introduces [e.g.  @Harji_2013].

Docker is limited to 64 bit host machines, making it impossible to run on
older hardware.

On Mac and Windows machines Docker must still be run in a fully
virtualized environment. Though the boot2docker tool streamlines this
process, it remains to be seen if the performance and integration with
the host machine's OS is sufficiently seamless or creates a barrier to
adoption by users on of these systems.

It remains to be seen if Docker will be significantly adopted by any scientific community.

@@

## Further considerations

<!-- Is this really necessary? -->

@@

<!-- Dynamic documents --> 
Dynamic documents are another example of an approach that to reproducible
research that can be facilitated through the use of Docker containers. 
Dynamic Documents [e.g. @Leisch_2002; @Peng_2011; @Xie_2013] seek to embed
the code required to re-generate the results within the manuscript
itself. This process serves as a way to distribute computational
code with publications, to ensure that methodological descriptions,
parameter values, figures, and so forth remain synchronized with the code,
and to provide a way for other researchers to both tweak the code and
'execute' the paper to reproduce the results.  While these features
are very desirable from the perspective of reproducible research,
they inevitably increase the number of software dependencies involved,
including those tools that support combining code with text [e.g. @Xie_2013]
as well as tools merely for the handling the formating and layout
(e.g. LaTeX, pandoc).  Consequently, such a paper can fail to execute
successfully for reasons having nothing to do with the computations
involved in the research, but rather due to software responsible for
implementing these dynamic features.  This problem is largely resolved by
providing a consistent development environment, making the approaches
discussed here a natural complement to that of dynamic documents.
A similar argument can be made regarding the use of workflow systems,
which can introduce additional software dependencies even as they seek
to make it easier to work across them.
